Monte Carlo solvers for sparse linear systems

### 2015-09-16 MAX_reports/MC_solvers_article/intro.tex
Editing work on mcls paper.

\section{Introduction}

@@

As underlined in \cite{DA1998}, Monte Carlo methods may be split into two
classes: \textit{direct methods} (\cite{DA1998}, \cite{DVA2001}) and
\textit{iterative methods} (\cite{Hal1962},
\cite{Hal1994}, \cite{ESW2013}
and \cite{EMSH2014}). The first are characterized by a merely stochastic
scheme,
therefore the provided error with respect to the exact solution is made of
just a stochastic component. The second ones, instead, combine numerical and
statistical schemes generating two types of error: a
\textit{stochastic} one and a \textit{systematic} one. It does not
mean
that
it will be always simple to recognize them separately. However it is important
to
be
aware of this intrinsic structure, in order to target what is the part of the
scheme that requires a refinement (e.g. increase of numerical iterations
rather than random walks).\newline

The paper is organized as follows. In Section 2 we describe all the
viable stochastic settings for the definition of the Markovian process. In
Section 3 we reinterpret probabilistic facts in terms of spectral properties of
the linear system to be solved. Section 4 is dedicated to necessary conditions
and sufficient conditions for the convergence of the algorithms. In Section 5
different adaptive approaches are presented, in order to truncate the random
walks and to decide how many samplings to use. Section 6 regards the study
of matrices for which the convergence of the stochastic schemes is facilitated.
Then in Section 7 a hybrid scheme combining numerical and probabilistic
techniques is described: the \textit{Monte Carlo Synthetic Acceleration}.
Numerical results are presented in Section 8 and 9, while Section 10 is
dedicated to
conclusive remarks and goals for the future.

>>>>

\section{Introduction}
\label{sec:intro}

@@

As underlined in \cite{DA1998}, Monte Carlo methods may be split into two
classes: \textit{direct methods} (\cite{DA1998}, \cite{DVA2001}) and
\textit{iterative methods} (\cite{Hal1962},
\cite{Hal1994}, \cite{ESW2013}
and \cite{EMSH2014}). The first are characterized by a merely stochastic
scheme,
therefore the provided error with respect to the exact solution is made of
just a stochastic component. The iterative Monte Carlo methods utilize more
traditional iterative algorithms alongside the stochastic approach,
generating two types of error: a
\textit{stochastic} one and a \textit{systematic} one. It does not
mean
that
it will be always simple to recognize them separately. However it is important
to
be
aware of this intrinsic structure, in order to target what is the part of the
scheme that requires a refinement (e.g. increasing the number of iterations
rather than the number of random walks).

The paper is organized as follows.
In Section~\ref{sec:mcls} we provide an overview of existing Monte Carlo
linear solver algorithms.
In Section~\ref{sec:convergence} we will discuss the convergence behavior
of stochastic solvers, including a discussion of classes of matrices for
which convergence can be guaranteed.
Section~\ref{sec:results} provides some numerical results illustrating
pertinent properties of the various approaches and Section~\ref{sec:conclusion}
will provide some concluding remarks and areas for future investigation.

### 2015-09-16 MAX_reports/MC_solvers_article/results.tex
Editing work on mcls paper.

\section{Numerical Results}

@@

The computation of the solution is realized both with the deterministic
Synthetic Acceleration and with Adjoint MCSA to compare the effectiveness of
the
stochastic technique. Results are reported in Table
\ref{Poisson_results}. For the sake of completeness, results obtained with
the Sequential Monte Carlo are shown as well. An evident discrepancy in terms
of numerical iterations
is already detected with the Sequential Monte Carlo outperforming the
preconditioned fixed point. Results are improved even more
when
switching to the MCSA, which employs less histories than the Sequential MC at
each iteration to
reach convergence. In fact the average number of histories employed by the
Sequential
Monte Carlo per iteration is almost one order of magnitude higher than for
MCSA. This
might be explained by the fact that the first row of the iterative algorithm in
the MCSA refines the accuracy of the solution coming from the previous
iteration,
before using it as the input for the MC solver at the current iterate.
In particular, since the first line of the algorithm is a deterministic step of
the Synthetic Acceleration, its effect is to dampen the statistical noise
produced by the Monte Carlo linear solver to estimate the update $\delta
x^{n + \frac{1}{2}}$ at the previous iteration. Therefore
refinement provided by the Synthetic Acceleration step decreases the number of
random walks
to employ for a prescribed accuracy. This is validated even more by the fact
that for the first iteration the Sequential Monte Carlo and the MCSA resort to
the same number of histories, while their behaviours start differing from the
second iteration on, when the statistical noise is introduced in the estimation
of the solution to \ref{linsys_prec} (see Figures \ref{SEQ_poisson} and
\ref{MCSA_poisson}).

>>>>

\section{Numerical Results}
\label{sec:results}

@@

The computation of the solution is realized both with the deterministic
Synthetic Acceleration and with Adjoint MCSA to compare the effectiveness of
the
stochastic technique. Results are reported in Table
\ref{Poisson_results}. For the sake of completeness, results obtained with
the Sequential Monte Carlo are shown as well. An evident discrepancy in terms
of numerical iterations
is already detected with the Sequential Monte Carlo outperforming the
preconditioned fixed point. Results are improved even more
when
switching to the MCSA, which employs less histories than the Sequential MC at
each iteration to
reach convergence. In fact the average number of histories employed by the
Sequential
Monte Carlo per iteration is almost one order of magnitude higher than for
MCSA. This
might be explained by the fact that the first row of the iterative algorithm in
the MCSA refines the accuracy of the solution coming from the previous
iteration,
before using it as the input for the MC solver at the current iterate.
In particular, since the first line of the algorithm is a deterministic step of
the Synthetic Acceleration, its effect is to dampen the statistical noise
produced by the Monte Carlo linear solver to estimate the update $\delta
x^{n + \frac{1}{2}}$ at the previous iteration. Therefore
refinement provided by the Synthetic Acceleration step decreases the number of
random walks
to employ for a prescribed accuracy. This is validated even more by the fact
that for the first iteration the Sequential Monte Carlo and the MCSA resort to
the same number of histories, while their behaviours start differing from the
second iteration on, when the statistical noise is introduced in the estimation
of the solution to \ref{linsys} (see Figures \ref{SEQ_poisson} and
\ref{MCSA_poisson}).
